Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures

### 2018-03-20 poster/qnmlPoster.tex
poster updated

\usepackage{textpos}
\usepackage{fancybox}
\usepackage{tikz}
\usepackage{mathtools}
\usepackage{empheq}

@@

   \begin{column}{0.3\textwidth} % second column {{{
     \heading{Structure Learning}
     \vspace*{12pt}
     \begin{itemize}
     \setlength\itemsep{1em}
     \item[Data:] Each $X_i$ is a categorical variable. We observe $n$ independent samples of $X$ which are collected in a data matrix $D$ of size $n\times d$. 
     \item[Goal:] We consider a \textit{score-based} approach and seek a graph $G$ that maximizes a scoring function which evaluates how well a given graph fits the observed data.
     \item Some scoring functions: BIC, BDeu and fNML. 
   	 \end{itemize}
   \end{column}

@@

\begin{block}{Quotient Normalized Maximum Likelihood Criterion}
\begin{columns}[T]
   \begin{column}{0.18\textwidth} % first column {{{
   \heading{Motivation}
   \vspace*{12pt}
   	\textcolor{teal}{BDeu}
	\begin{itemize}	    
    \item Is very sensitive to the choice of the hyperparameter $\alpha$ \cite{cosco.uai07}.
    \item Is not \textit{regular} \cite{Suzuki2017} (can be shown in certain situations to favour too complex models over simpler) 
    \end{itemize}
    \vspace*{12pt}
    \textcolor{teal}{fNML}
	\begin{itemize}	    
    \item Is not score equivalent: the graphs expressing the same independence statements are not scored equally.
    \item Learned structures are often rather complex, which hampers their interpretation.
    \end{itemize}

    %\vspace*{12pt} 
	%\textcolor{HIITgreen}{Figure:} Number of parameters as a function of sample size for UCI Breast Cancer  data.

@@

   \begin{column}{0.18\textwidth}
   \vphantom{\large Motivation}
   \vspace*{12pt} 
    \textcolor{teal}{BIC}
    \begin{itemize}
    \item Appears to
require large sample sizes in order to identify appropriate
structures \cite{cosco.pgm08a,Liu2012}. 
	\end{itemize}

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){650}
   \end{column} % }}}

@@

	\item However, for \textbf{a single categorical data vector}, regret and one dimensional $P^1_{NML}$-code can be computed.
	%\item In $P^1_{NML}$, regret depends on $N$, the length of the categorical data vector $D_i$ and $r_i$, the number of possible values for corresponding variable $X_i$. 
	%\item Justification from \textit{Minimum description length} principle.
	\item \textbf{Trick}. Recall factorization w.r.t. graph $G$:
	$$
	P(D\mid G) = \prod_{i = 1}^{d}P(D_i \mid D_{G_i}) = \prod_{i = 1}^{d}\frac{P(D_{i,G_i})}{P(D_{G_i})}
	$$
	\item Assume there are \textbf{no independencies} among $G_i$. Now, $D_{i,G_i} = $
	\vspace*{12pt}
	\end{itemize}
	\begin{minipage}{0.45\textwidth}
	\begin{align*}
	\begin{blockarray}{c|ccc}
	D_i\phantom{s} & \phantom{..}  & D_{G_i} & \phantom{..}\\
	\begin{block}{(c|ccc)}
  	1 & 1 & 1 & 1  \\
  	0 & 1 & 0 & 0 \\
  	1 & 1 &0 & 0  \\
  	1 & 1 & 1 & 1  \\
	\end{block}
	\end{blockarray} 
	\ \to \ 
	\begin{blockarray}{c}
	\phantom{...}\\
	\begin{block}{(c)}
  	0   \\
  	1  \\
  	2  \\
  	0   \\
	\end{block}
	\end{blockarray}
 	\end{align*}
	\end{minipage}\quad
	\begin{minipage}{0.45\textwidth}
	\fbox{\parbox{\textwidth}{
	\begin{itemize}
	\item We can treat $D_{i,G_i}$ and $D_{G_i}$ as values of single variables!
	\item $P(D_{i,G_i})$ and $P(D_{i,G_i})$ can be modeled using $P_{NML}^1$-code.
	\end{itemize}}}  
	\end{minipage}

@@

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){650}
   \end{column} % }}}

@@

   	\heading{qNML: Definition \& Properties}
   	Definition of qNML score:
    	\begin{empheq}[box={\mybluebox[4pt][4pt]}]{equation}	    		
	s^{qNML} := \sum_{i = 1}^d \log \frac{P^1_{NML}(D_{i,G_i})}{P^1_{NML}(D_{G_i})},\notag
	\end{empheq}$\log P^1_{NML}(D) = \log P(D \mid \hat{\theta}) - reg(n,r).$

	\begin{itemize}
	 \item $D$ is understood as  $n$ observations on single categorical variable with $r$ possible values.
	\item Accurate approximation for $reg(n,r)$ from \cite{Szpankowski2012}:
	\begin{align*}
    reg(n, r) &\approx  n\left(\log{\alpha} + (\alpha + 2) \log{C_\alpha} - \frac{1}{C_\alpha}\right) \\ 
    &- \frac{1}{2} \log{\left(C_\alpha + \frac{2}{\alpha}\right)},
	\end{align*}
where $\alpha = \frac{r}{n}$ and $C_\alpha = \frac{1}{2} + \frac{1}{2} \sqrt{1 + \frac{4}{\alpha}}$.
	\end{itemize}
	\vspace*{12pt}
	qNML score is:
	\fbox{\parbox{0.95\textwidth}{
	\begin{itemize}	
	\item \textcolor{teal}{\textbf{hyper-parameter free}}
	\item \textcolor{teal}{\textbf{consistent}}
	\item \textcolor{teal}{\textbf{score equivalent}}
	\item \textcolor{teal}{\textbf{regular}}
	\item qNML agrees with NML for every network composed of fully connected sub-graphs.
	\end{itemize}}}

@@

	\vspace*{12pt}
	\textbf{qNML seems overall a safe choice in structure learning, especially with moderate and large sample sizes!} 
    \end{column}
   \begin{column}{0.005\textwidth}\linethickness{0.3ex}
      \color{myPurple} \line(0,1){600}
   \end{column} % }}}

@@

    \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){600}
   \end{column} % }}}
    \begin{column}{0.3\textwidth}
    \heading{Conclusions}

>>>>

\usepackage{textpos}
\usepackage{fancybox}
\setlength{\fboxsep}{12pt}
%\usepackage{mathtools}
\usepackage{empheq}

@@

   \begin{column}{0.25\textwidth} % second column {{{
     \heading{Structure Learning}
     \vspace*{12pt}
     \begin{itemize}
     \setlength\itemsep{1em}
     \item[Data:] Each $X_i$ is a categorical variable. We observe $n$ independent samples of $X$ which are collected in a data matrix $D$ of size $n\times d$. 
     \item[Goal:] We consider a \textit{score-based} approach and seek a graph $G$ that maximizes a scoring function which evaluates how well a given graph fits the observed data.
     \item Some scoring functions: BIC, BDeu and fNML. 
   	 \end{itemize}
   \end{column}

@@

\begin{block}{Quotient Normalized Maximum Likelihood Criterion}
\begin{columns}[T]

	\begin{column}{0.34\textwidth} % first column {{{
   		\heading{Motivation}
   		\begin{minipage}{0.49\textwidth}
   			\vspace*{12pt}
		   	\textcolor{teal}{BDeu}
			\begin{itemize}	    
		    \item Is very sensitive to the choice of the hyperparameter $\alpha$ \cite{cosco.uai07}.
		    \item Is not \textit{regular} \cite{Suzuki2017} (can be shown in certain situations to favour too complex models over simpler) 
		    \end{itemize}
		    \vspace*{12pt}
		    \textcolor{teal}{fNML}
			\begin{itemize}	    
		    \item Is not score equivalent: the graphs expressing the same independence statements are not scored equally.
		    \item Learned structures are often rather complex, which hampers their interpretation.
		    \end{itemize}

   		\end{minipage}
   		\begin{minipage}{0.49\textwidth}		   
		   \phantom{why vspace* does nothing??}
		   %\vspace*{12pt} 
		   \textcolor{teal}{BIC}
		    \begin{itemize}
		    \item Appears to
		require large sample sizes in order to identify appropriate
		structures \cite{cosco.pgm08a,Liu2012}. 
			\end{itemize}

   		\end{minipage}

   		\vspace*{12pt}

   		\shadowbox{
		\parbox{0.97\textwidth}{
		\vspace*{12pt}
		The quest for a model selection criterion that would yield \textbf{more parsimonious}, \textbf{easier to interpret}, but still \textbf{predictive} Bayesian networks structures is one of the main motivations for this work.
		\vspace*{12pt}
		}}

@@

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){810}
   \end{column} % }}}

@@

	\item However, for \textbf{a single categorical data vector}, regret and one dimensional $P^1_{NML}$-code can be computed.
	%\item In $P^1_{NML}$, regret depends on $N$, the length of the categorical data vector $D_i$ and $r_i$, the number of possible values for corresponding variable $X_i$. 
	%\item Justification from \textit{Minimum description length} principle.
	\item \textbf{Trick}. Recall factorization w.r.t. graph $G$:
	$$
	P(D\mid G) = \prod_{i = 1}^{d}P(D_i \mid D_{G_i}) = \prod_{i = 1}^{d}\frac{P(D_{i,G_i})}{P(D_{G_i})}
	$$
	\item Assume there are \textbf{no independencies} among $G_i$. Now, $D_{i,G_i} = $
	\vspace*{12pt}
	\end{itemize}
	\begin{minipage}{0.45\textwidth}
	\begin{align*}
	\begin{blockarray}{c|ccc}
	D_i\phantom{s} & \phantom{..}  & D_{G_i} & \phantom{..}\\
	\begin{block}{(c|ccc)}
  	1 & 1 & 1 & 1  \\
  	0 & 1 & 0 & 0 \\
  	1 & 1 &0 & 0  \\
  	1 & 1 & 1 & 1  \\
	\end{block}
	\end{blockarray} 
	\ \to \ 
	\begin{blockarray}{c}
	\phantom{...}\\
	\begin{block}{(c)}
  	0   \\
  	1  \\
  	2  \\
  	0   \\
	\end{block}
	\end{blockarray}
 	\end{align*}
	\end{minipage}\quad
	\begin{minipage}{0.45\textwidth}
	\shadowbox{
	\parbox{\textwidth}{
	\vspace*{12pt}
	\begin{itemize}
	\item We can treat $D_{i,G_i}$ and $D_{G_i}$ as values of single variables!
	\item $P(D_{i,G_i})$ and $P(D_{i,G_i})$ can be modeled using $P_{NML}^1$-code.
	\end{itemize}}}  
	\end{minipage}

@@

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){810}
   \end{column} % }}}

@@

   	\heading{qNML: Definition \& Properties}
   	\vspace*{12pt}
   	Definition of qNML score:
    	\begin{empheq}[box={\mybluebox[4pt][4pt]}]{equation}	    		
	s^{qNML} := \sum_{i = 1}^d \log \frac{P^1_{NML}(D_{i,G_i})}{P^1_{NML}(D_{G_i})},\notag
	\end{empheq}$\log P^1_{NML}(D) = \log P(D \mid \hat{\theta}) - reg(n,r).$

	\begin{itemize}
	 \item $D$ is understood as  $n$ observations on single categorical variable with $r$ possible values.
	\item Accurate approximation for $reg(n,r)$ from \cite{Szpankowski2012}:
	\begin{align*}
    reg(n, r) &\approx  n\left(\log{\alpha} + (\alpha + 2) \log{C_\alpha} - \frac{1}{C_\alpha}\right) \\ 
    &- \frac{1}{2} \log{\left(C_\alpha + \frac{2}{\alpha}\right)},
	\end{align*}
where $\alpha = \frac{r}{n}$ and $C_\alpha = \frac{1}{2} + \frac{1}{2} \sqrt{1 + \frac{4}{\alpha}}$.
	\end{itemize}

	\vspace*{12pt}
	\shadowbox{\parbox{0.85\textwidth}{
	\vspace*{12pt}
	\textbf{qNML score is:}
	\begin{itemize}	
	\item \textcolor{teal}{\textbf{hyper-parameter free}}
	\item \textcolor{teal}{\textbf{consistent}}
	\item \textcolor{teal}{\textbf{score equivalent}}
	\item \textcolor{teal}{\textbf{regular}}
	\item \textcolor{teal}{\textbf{it equals NML for many models.}}
	%\item qNML agrees with NML for every network composed of fully connected sub-graphs.
	\end{itemize}}}

@@

	\vspace*{12pt}
	%\textbf{qNML seems overall a safe choice in structure learning, especially with moderate and large sample sizes!} 
    \end{column}
   \begin{column}{0.005\textwidth}\linethickness{0.3ex}
      \color{myPurple} \line(0,1){590}
   \end{column} % }}}

@@

    \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){590}
   \end{column} % }}}
    \begin{column}{0.3\textwidth}
    \heading{Conclusions}
